---
title: "p8105_group_proposal"
author: "NSK, ST (sjt2164), Shayne Estill(ske2118), WS"
date: "2024-11-06"
output: github_document
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)

```

## R Markdown

```{r}


```

#### Tentative Project Title  

The impact of NYC Environmental Factors on Children’s Health  


#### Motivation for this project 

Children’s health can be impacted by their surrounding environment. Poor air quality due to air pollution, especially in urban environments, can influence respiratory health and have long-term consequences. The availability of green spaces has been found to have an association with physical, mental, and social well-being for child development. Thus, a lack of green spaces, in combination with urban air pollution could have negative impacts. With the increasing concerns of climate change, the heat island effect in cities could also lead to an increase in heat-related illness. Our main project aim is to investigate the effect of environmental factors on respiratory health across NYC boroughs. We are also interested in looking into how green space availability and heat can play a role. 

 
