---
title: "p8105_group_proposal"
author: "NSK, ST (sjt2164), Shayne Estill(ske2118), WS"
date: "2024-11-06"
output: github_document
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)

```

## R Markdown

```{r}


```

#### Tentative Project Title  

The impact of NYC Environmental Determinants on Children’s Health  

#### Motivation for this project 

Children’s health and development is impacted by the surrounding environment, particularly in dense urban environments like New York City (NYC). Poor air quality due to urban air pollution can influence and have long-term consequences for child and adolescent respiratory health. The challenges of urban environmental health are further complicated by climate change, particularly through the urban heat island effect, which can increase the risk of heat-related illnesses. Limited research shows that elevated temperatures due to the urban heat island effect are associated with increased emergency department (ED) visits among children (Sheffield et al., 2018). These health outcomes may be influenced by the availability of green spaces, which can help filter air pollutants and provide cooler areas for shelter from heat. Importantly, environmental health risks disproportionately affect children from low-income neighborhoods and communities of color, where there is often limited access to green spaces and higher exposure to environmental stressors. 

Through this project, we aim to investigate how these interconnected environmental determinants- air quality, green space availability, and urban heat patterns - vary across NYC boroughs and their potential associations with children’s health, specifically focusing on respiratory outcomes and ED visits. We believe that understanding these relationships is crucial for informing urban planning policies and public health initiatives aimed at improving children’s health in New York City. 

 
#### Anticipated Data Sources

* Environmental data:  

** New York City Air Quality (Kaggle: https://www.kaggle.com/datasets/sahityasetu/new-york-city-air-quality) 

** NOAA Weather Data for temperature (https://www.ncei.noaa.gov/metadata/geoportal/rest/metadata/item/gov.noaa.ncdc:C00861/html)  

** NYC Parks and Recreation Data (https://www.nycgovparks.org/about/data)  

* Health and demographic data:  

** For ED visits and respiratory health outcomes (https://a816-health.nyc.gov/hdi/epiquery/visualizations?PageType=ts&PopulationSource=Syndromic&Topic=1&Subtopic=39)  