---
title: "p8105_group_final"
output: github_document
author: Shayne Estill (ske2118) , WooJin Shin (ws2721), Naomi Simon-Kumar (ns3782), Serena Ting (sjt2164)
---

* Detail how you completed your project
* Cover data collection and cleaning, exploratory analyses, alternative strategies, descriptions of approaches, and a discussion of result
* Document any changes and the reasons 
* One report document per group (include all group member names)

## Motivation: 

Provide an overview of the project goals and motivation.

## Related work: 

Anything that inspired you, such as a paper, a web site, or something we discussed in class.

## Initial questions: 

What questions are you trying to answer? How did these questions evolve over the course of the project? What new questions did you consider in the course of your analysis?

## Data: 

Source, scraping method, cleaning, etc.

## Exploratory analysis: 

Visualizations, summaries, and exploratory statistical analyses. Justify the steps you took, and show any major changes to your ideas.

## Additional analysis: 
If you undertake formal statistical analyses, describe these in detail


## Discussion: 

What were your findings? Are they what you expect? What insights into the data can you make?

As this will be your only chance to describe your project in detail, make sure that your report is a standalone document that fully describes your process and results.

We also expect you to write high-quality code that is understandable to an outside reader. Coding collaboratively and actively reviewing code within the team will help with this!

------------------

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)

```


## P8105: Group Proposal Submission

Shayne Estill (ske2118) , WooJin Shin (ws2721), Naomi Simon-Kumar (ns3782), Serena Ting (sjt2164)

#### Tentative Project Title  

The impact of NYC Environmental Determinants on Children’s Health  

#### Motivation for this project 

Children’s health and development is impacted by the surrounding environment, particularly in dense urban environments like New York City (NYC). Poor air quality due to urban air pollution can influence and have long-term consequences for child and adolescent respiratory health. The challenges of urban environmental health are further complicated by climate change, particularly through the urban heat island effect, which can increase the risk of heat-related illnesses (Heaviside et al., 2017). Limited research shows that elevated temperatures due to the urban heat island effect are associated with increased emergency department (ED) visits among children (Sheffield et al., 2018). These health outcomes may be influenced by the availability of green spaces, which can help filter air pollutants and provide cooler areas for shelter from heat. Importantly, environmental health risks disproportionately affect children from low-income neighborhoods and communities of color, where there is often limited access to green spaces and higher exposure to environmental stressors. 

Through this project, we aim to investigate how these interconnected environmental determinants- air quality, green space availability, and urban heat patterns - vary across NYC boroughs and their potential associations with children’s health, specifically focusing on respiratory outcomes and ED visits. We believe that understanding these relationships is crucial for informing urban planning policies and public health initiatives aimed at improving children’s health in New York City. 

 
#### Anticipated Data Sources

* Environmental data:  

New York City Air Quality (Kaggle: https://www.kaggle.com/datasets/sahityasetu/new-york-city-air-quality) 

NOAA Weather Data for temperature (https://www.ncei.noaa.gov/metadata/geoportal/rest/metadata/item/gov.noaa.ncdc:C00861/html)  

NYC Parks and Recreation Data (https://www.nycgovparks.org/about/data)  

* Health and demographic data:  

For ED visits and respiratory health outcomes (https://a816-health.nyc.gov/hdi/epiquery/visualizations?PageType=ts&PopulationSource=Syndromic&Topic=1&Subtopic=39)  

#### Intended final product

The intended final project will be a form of an interactive webpage presenting our research findings, summarizing the associations between environmental factors and pediatric health outcomes across NYC boroughs. We intend to include the following sections:


1. Interactive Visualizations:  

* Air pollution distribution maps by borough 

* Urban heat island effect maps by borough 

* Green space accessibility visualization 

* Comparative visualizations of environmental determinants and demographic characteristics by borough 

2. Statistical Analysis:  

* Linear regression models examining associations between:  

* Air quality and respiratory health outcomes stratified by demographic groups 

* Temperature patterns and ED visits stratified by demographic groups 

* Analysis of effects of green space availability 

3. Discussion of Future Directions for research and practice - in the form of brief recommendations for city planners and public health officials

#### The planned analysis / visualization / coding challenges

There may be some challenges with missing data and merging data between different data sources, particularly when working collaboratively and in parallel with our team. The main analytical challenge will be building appropriate linear models to explore the effects of multiple environmental variables (air quality, temperature, green space) on health outcomes with the data available to us, while accounting for relevant demographic factors.  

Our planned analyses: 

1. Data Wrangling: Clean, tidy, and merge environmental, health, and demographic datasets 

2. Visualization (exploratory analysis) 

   a. Descriptive data visualization using basic plots 

   b. Simple maps using borough boundaries 

   c. Paneled plots to compare boroughs 

   d. Interactive dashboard using flexdashboard 

3. Statistical Analysis  

    Linear models to test relationships between:  

   a. Air quality and respiratory health outcomes 

   b. Temperature and ED visits

   c. Effects of greenspace  
      

#### The planned timeline

November 11-15 tidy the data for analysis 

November 18-22 Finish the analysis and write the report 

November 25-29 finish the report and build a webpage     


~ December 6th Finish the rest (screencast, Peer assessment)  

#### References 

Heaviside, C., Macintyre, H., & Vardoulakis, S. (2017). The urban heat island: implications for health in a changing environment. Current environmental health reports, 4, 296-305.

Sheffield, P. E., Herrera, M. T., Kinnee, E. J., & Clougherty, J. E. (2018). Not so little differences: variation in hot weather risk to young children in New York City. Public health, 161, 119–126. https://doi.org/10.1016/j.puhe.2018.06.004 



##### Data Cleaning: Pediatric ED Admissions and Hospitalizations data

```{r}

# Load libraries 

library(tidyverse)

```

```{r}

# Load and clean data for 2016-2018

ed_2016_2018 <- read_csv("data/ed_data_csv.csv", skip = 2) %>%
 janitor::clean_names() %>%
 filter(
   data_years == "2016-2018",
   indicator %in% c(
     "Asthma emergency department visit rate per 10,000 - aged 0-17 years",
     "Asthma hospitalization rate per 10,000 - aged 0-17 years"
   ),
   county %in% c("New York", "Queens", "Richmond", "Kings", "Bronx")
 ) %>%
 select(data_years, county, zip_code, indicator, zip_code_rate) %>%
 mutate(
   indicator = case_when(
     indicator == "Asthma emergency department visit rate per 10,000 - aged 0-17 years" ~ "ed_visits_rate",
     indicator == "Asthma hospitalization rate per 10,000 - aged 0-17 years" ~ "hosp_rate"
   ),
   zip_code_rate = case_when(
     str_detect(zip_code_rate, "\\*") ~ NA_character_,
     zip_code_rate == "s" ~ NA_character_,
     TRUE ~ zip_code_rate
   )
 ) %>%
  rename(borough = county) %>%
  mutate(
    borough = case_when(
      borough == "New York" ~ "Manhattan",
      borough == "Kings" ~ "Brooklyn",
      borough == "Richmond" ~ "Staten Island",
      TRUE ~ borough
    )
  )

# Load and clean data for 2019-2021

ed_2019_2021 <- read_csv("data/ed_data_csv.csv", skip = 2) %>%
 janitor::clean_names() %>%
 filter(
   data_years == "2019-2021",
   indicator %in% c(
     "Asthma emergency department visit rate per 10,000 - aged 0-17 years",
     "Asthma hospitalization rate per 10,000 - aged 0-17 years"
   ),
   county %in% c("New York", "Queens", "Richmond", "Kings", "Bronx")
 ) %>%
 select(data_years, county, zip_code, indicator, zip_code_rate) %>%
 mutate(
   indicator = case_when(
     indicator == "Asthma emergency department visit rate per 10,000 - aged 0-17 years" ~ "ed_visits_rate",
     indicator == "Asthma hospitalization rate per 10,000 - aged 0-17 years" ~ "hosp_rate"
   ),
   zip_code_rate = case_when(
     str_detect(zip_code_rate, "\\*") ~ NA_character_,
     zip_code_rate == "s" ~ NA_character_,
     TRUE ~ zip_code_rate
   )
 ) %>%
  rename(borough = county) %>%
  mutate(
    borough = case_when(
      borough == "New York" ~ "Manhattan",
      borough == "Kings" ~ "Brooklyn",
      borough == "Richmond" ~ "Staten Island",
      TRUE ~ borough
    )
  )
```

The data cleaning process involved multiple steps to prepare asthma-related health data across two time periods (2016-2018 and 2019-2021) for analysis. Data was obtained from the New York State Asthma Dashboard, which compiles information from multiple sources including the Behavioral Risk Factor Surveillance System (BRFSS) and Statewide Planning and Research Cooperative System (SPARCS) databases. For each time period, data was loaded from a CSV file, skipping the first two rows, and variable names were standardized using the janitor package's clean_names() function to ensure consistent formatting.

The data was then filtered to include two specific health indicators: asthma emergency department visit rates and asthma hospitalization rates for ages 0-17 years. The geographic scope was limited to five counties (New York, Queens, Richmond, Kings, and Bronx), and only relevant variables were retained. These variables included data_years (the specified three-year period of data collection), county (geographic location), zip_code (postal code), indicator (type of healthcare utilization), and zip_code_rate (rate per 10,000 children aged 0-17 years).

Variables were renamed: "Asthma emergency department visit rate per 10,000 - aged 0-17 years" to "ed_visits_rate" and "Asthma hospitalization rate per 10,000 - aged 0-17 years" to "hosp_rate". County was renamed to borough to reflect NYC terminology. This involved changing "New York" to "Manhattan", "Kings" to "Brooklyn", and "Richmond" to "Staten Island". 

Missing and unstable values in zip_code_rate were handled in two ways: values marked with "s" were converted to NA (indicating data suppressed for confidentiality reasons due to not meeting reporting criteria), and values marked with an asterisk were converted to NA (indicating fewer than 10 events in the numerator, making the rate unstable).

The resulting datasets provide clean, consistently formatted data for analyzing pediatric asthma-related healthcare utilization across New York City boroughs at the ZIP code level for 2016-2018 and 2019-2021.

##### Temperature

https://open-meteo.com/en/docs/historical-weather-api

used this website and recorded the coordinates of 5 boroughs and extracted the temperatrue information by daily in a timeframe between 2016 and 2021.


```{r}

manhattan = read_csv("data/manhattan.csv",skip = 2) |>
  mutate(indicator = "manhattan") |> 
  janitor::clean_names()

bronx = read_csv("data/bronx.csv",skip = 2) |>
  mutate(indicator = "bronx") |> 
  janitor::clean_names()

brooklyn = read_csv("data/brooklyn.csv",skip = 2) |>
  mutate(indicator = "brooklyn") |> 
  janitor::clean_names()

queens = read_csv("data/queens.csv",skip = 2) |>
  mutate(indicator = "queens") |> 
  janitor::clean_names()

staten_island = read_csv("data/staten_island.csv",skip = 2) |>
  mutate(indicator = "staten_island") |> 
  janitor::clean_names()
```

I then merged the data by borough and divided into two groups to compare.

```{r}
all_temp = bind_rows(manhattan, bronx, brooklyn, queens, staten_island) |> 
  mutate(indicator = as.factor(indicator)) |>
  mutate(date = as.Date(time, format = "%Y-%m-%d")) |> 
  mutate(date_group = ifelse(year(date) %in% 2016:2018, "2016-2018", "2019-2021")) |> 
  select(temperature_2m_max_c:date_group) |> 
  group_by(date_group, indicator) |>
  summarise(avg_temperature = mean(temperature_2m_mean_c, na.rm = TRUE))
  

write.csv(all_temp, "data/NYC_temp.csv", row.names = FALSE)

all_temp_raw = bind_rows(manhattan, bronx, brooklyn, queens, staten_island) |> 
  mutate(indicator = as.factor(indicator))

write.csv(all_temp_raw, "data/NYC_temp_raw.csv", row.names = FALSE)

ggplot(data = all_temp, aes(x = date_group, y = avg_temperature, color = indicator)) +
  geom_point(size = 3) +
  geom_line(aes(group = indicator), size = 1) +
  labs(title = "Average Temperature by Date Group and Indicator",
       x = "Date Group",
       y = "Average Temperature") +
  theme_minimal()

```

## Greenspace Data

Greenspace data: https://data.cityofnewyork.us/Recreation/Parks-Properties/enfh-gkve/about_data

```{r greenspace, include=FALSE}
greenspace_df = read_csv("Data/Parks_Properties_20241126.csv", na = c("NA", ".", "")) |>
  janitor::clean_names()
```

```{r}
greenspace_clean = greenspace_df |>
  mutate(
    borough = case_match(borough, 
      "R" ~ "Staten Island",
      "Q" ~ "Queens",
      "X" ~ "Bronx",
      "B" ~ "Brooklyn",
      "M" ~ "Manhattan"),
    typecategory = as.factor(typecategory),
    borough = as.factor(borough)) |>
  separate(acquisitiondate, into = c("year", "month", "day"), sep = "-") |>
  select(year, borough, acres) |> 
  filter(year < 2022 | is.na(year))
   
#keep date NA; up to 2019

summary(greenspace_clean)
```


#### Dataset 1: 2016-2018 (3 years)

```{r}
dataset1_gs = greenspace_clean |>
  filter(year < 2019 | is.na(year)) |> 
  group_by(year, borough) |>
  summarise(avg_acres_per_yr_bor = mean(acres, na.rm = TRUE), .groups = "drop")

dataset1_gs_calc = dataset1_gs |> 
  group_by(borough) |> 
  summarise(acres_sum = sum(avg_acres_per_yr_bor, na.rm = TRUE), .groups = "drop") |> 
  mutate(year_group = "2016-2018")
```


#### Dataset 2: 2019-2021 (3 years)

```{r}
dataset2_gs = greenspace_clean |>
  filter(year < 2022 | is.na(year)) |>
  group_by(year, borough) |>
  summarise(avg_acres_per_yr_bor = mean(acres, na.rm = TRUE), .groups = "drop")
<<<<<<< HEAD

dataset2_gs_calc = dataset2_gs |> 
  group_by(borough) |> 
  summarise(acres_sum = sum(avg_acres_per_yr_bor, na.rm = TRUE), .groups = "drop") |> 
  mutate(year_group = "2019-2021")
```

#### comparing by the year

```{r}
total_acres = bind_rows(dataset1_gs_calc, dataset2_gs_calc) |>
  mutate(
    percentage = case_when(
      borough == "Manhattan" ~ (acres_sum/300800)*100,
      borough == "Bronx" ~ (acres_sum/571072)*100,
      borough == "Brooklyn" ~ (acres_sum/1093312)*100,
      borough == "Queens" ~ (acres_sum/1349120)*100,
      borough == "Staten Island" ~ (acres_sum/1127936)*100
    )
  ) |> 
  select(borough, percentage, year_group) |> 
  pivot_wider(names_from = year_group,
              values_from = percentage)

total_acres
```

=======
```

### PM2.5


```{r}
library(tidyverse)
library(rvest)
library(readxl)
```


Import and clean PM2.5 data from website: https://www.kaggle.com/datasets/sahityasetu/new-york-city-air-quality?resource=download. 


```{r}
particulate_matter = 
  read_csv("data/Air_Quality_20231208.csv")
```

The raw data has `r nrow(particulate_matter)` rows and `r ncol(particulate_matter)` columns. 

The columns are Unique ID, Indicator ID, Name, Measure, Measure Info, Geo 
Type Name, Geo Join ID, Geo Place Name, Time Period, Start_Date, Data Value, and Message. 

Unique ID: A unique identifier assigned to each row in the dataset.

Indicator ID: A code assigned to each indicator or measure of air quality being tracked.

Name: The name or label given to the indicator or measure being tracked.

Measure: The unit of measurement used to quantify the air quality indicator, such as parts per billion (ppb) for ozone or sulfur dioxide.

Measure Info: Additional information about the measurement or calculation of the air quality indicator, if applicable.

Geo Type Name: The type of geographic area being tracked, such as community districts (CDs) or boroughs.

Geo Join ID: A unique identifier assigned to each geographic area being tracked.

Geo Place Name: The name of the specific geographic area being tracked, such as Coney Island or the Bronx.

Time Period: The time period during which the air quality measurement was taken, such as a specific season or winter of a particular year.

Start_Date: The date on which the air quality measurement period began.

Data Value: The value of the air quality indicator for the specific geographic area and time period being tracked.

Message: Additional information or notes about the air quality measurement or data value, if applicable.

As it stands, these data are not "tidy": Unique ID should be a character, as should Indicator ID and Geo Join ID. 

We are interested in PM2.5 so we can filter name to "Fine particles (PM 2.5)". I should note there is name option "Asthma emergency department visits due to PM2.5", "Respiratory hospitalizations due to PM2.5 (age 20+)", "Cardiovascular hospitalizations due to PM2.5 (age 40+)", "Deaths due to PM2.5", "Respiratory hospitalizations due to PM2.5 (age 20+)", corresponding to annual rates(18+, 30+, etc.) that could be interesting to look at in secondary analyses. 

Further filter to years of interest, we will be looking to compare 2016-2018 to 2019-2021.

We will make two datasets, one for analysis in which we will filter by borough and create a new column that has the average PM2.5 across 2016-2018 and 2019-2021. 
```{r}
tidy_analysis_pm = 
  particulate_matter |>
  janitor::clean_names() |>
    mutate(
    unique_id = as.character(unique_id),
    indicator_id = as.character(indicator_id),
    geo_join_id = as.character(geo_join_id)) |>
  filter(name == "Fine particles (PM 2.5)") |>
  filter(time_period %in% c("Annual Average 2016", "Annual Average 2017",
                            "Annual Average 2018", "Annual Average 2019",
                            "Annual Average 2020", "Annual Average 2021")) |>
  filter(geo_type_name == "Borough") |>
  select(-unique_id, -indicator_id, -start_date) |>
    rename(borough = geo_place_name) |>
  mutate(
    period = case_when(
      time_period %in% c("Annual Average 2016", "Annual Average 2017", "Annual Average 2018") ~ "2016-2018",
      time_period %in% c("Annual Average 2019", "Annual Average 2020", "Annual Average 2021") ~ "2019-2021"
    )
  ) |>
  group_by(borough, period) |>
  summarize(
    average_pm2.5 = mean(data_value, na.rm = TRUE),
    .groups = "drop"
  ) |>
  pivot_wider(
    names_from = period,
    values_from = average_pm2.5,
    names_prefix = "avg_pm2.5_"
  )
```

Cleaned analysis PM2.5 dataset has `r nrow(tidy_analysis_pm)` rows and `r ncol(tidy_analysis_pm)` columns. 

I made a new dataset that is by borough and individual years to be used in the visualization step. 

```{r}
tidy_visualization_pm = 
  particulate_matter |>
  janitor::clean_names() |>
    mutate(
    unique_id = as.character(unique_id),
    indicator_id = as.character(indicator_id),
    geo_join_id = as.character(geo_join_id)) |>
  filter(name == "Fine particles (PM 2.5)") |>
  filter(time_period %in% c("Annual Average 2016", "Annual Average 2017",
                            "Annual Average 2018", "Annual Average 2019",
                            "Annual Average 2020", "Annual Average 2021")) |>
  filter(geo_type_name == "CD") |>
    rename(pm2.5_mcgm3 = data_value) 
```


```{r}
geography = 
  read_excel("data/geoid_borough_name_nyc.xlsx") |>
  rename(geo_join_id = ID) |>
  mutate(geo_join_id = 
           as.character(geo_join_id))
```


```{r}
pm_for_vis = 
    left_join(tidy_visualization_pm, geography, by = c("geo_join_id")) |>
    rename(neighborhood = Name) |>
    rename(borough = Borough) |>
   select(name, neighborhood, borough, time_period, pm2.5_mcgm3)
```

Cleaned visualization PM2.5 dataset has `r nrow(pm_for_vis)` rows and `r ncol(pm_for_vis)` columns. 

This merged on the geo_join_id column for each community district. I renamed some variables for clarity and selected the variables of interest for visualization.
>>>>>>> fa1dce32d0a33e55330f43be66c8e85822117900
